Drug sensitivity prediction with the normal inverse Gaussian prior informed by external data

### 2019-11-20 docs/manuscript1.Rnw
created different selection strategies in analysis_gdsc.R

	The GDSC database contains additional information on both the drugs and 
	molecular features, such as the target 
	pathways and developmental stages of the drugs. Additional online
	repositories might provide extra information such as the molecular weight of
	the compounds or the publication signatures of the molecular features. In some
	cases, prior knowledge on the drug efficacies might be available,
	from previous experiments. We propose to include these possibly beneficial 
	information sources in the estimation of the sensitivity prediction models in 
	a data-driven manner. More specifically, we estimate a normal 
	inverse Gaussian model, where the amount of regularisation is estimated by an 
	adaptive empirical Bayes procedure, guided by the external information.

@@

	The rest of the paper is structured as follows. In Section 
	\ref{sec:model} we introduce our model, the estimation of which is detailed
	in Section \ref{sec:estimation}. In Section \ref{sec:gdsc} we analyse the
	GDSC data, and we end with a discussion in Section \ref{sec:discussion}.

@@

	\subsection{Bayesian prior model}
	We carry out inference by endowing the parameters with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
		  \label{eq:tauprior}\\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\phi$ and shape $\lambda$. Prior distributions of the form
	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
	prior 
	variance into a local components $\gamma_{jd}^2$ and a global component
	$\tau_d^2$. The local components are supposed to capture local variation in
	the model parameters $\beta_{jd}$, while the global component is supposed to
	capture the general trend in $\beta_{jd}$. Addionally, for appropriate local
	shrinkage in global-local 
	shrinkage models it is important  to account for 
  different noise levels $\sigma_d^2$ by scaling the $\beta_{jd}$ variances 
  accordingly.

	The normal inverse Gaussian (NIG) prior model was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and since 
  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
  additional global variance component $\tau_d^2$. Supplementary Material
  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
  To illustrate the effect of the NIG prior on the posterior mean, we consider 
  the prior reparametrised as in 
  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
  Under the (simplified) normal means model, i.e., 
  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
  $\tau_d^2=\sigma_d^2=1$, the resulting
  posterior mean for the $\beta_{jd}$ is
  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
  implies full shrinkage towards zero. Figure 
  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
  several choices of $\beta_{jd}$ prior.
<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
@

@@

	\subsection{External information}\label{sec:external}
	In drug sensitivity prediction models external information on both the drugs
	and features is often available. Here, we assume this information to be 
	available as
	external feature `covariates' $\mathbf{c}_{jdg}$, for $g=1, \dots, G$, and 
	drug `covariates' $\mathbf{z}_{dh}$, for $h=1,\dots, H$. An example of a 
	(binary) features covariate is target pathway presence, with $c_{jdg}=0$ if 
	gene $j$ is present in the 
	target pathway of drug $d$ and $c_{jdg}=1$ if it is not. An example of a 
	(ternary) drug
	covariate is developmental phase, with levels experimental phase, 
	clinical development, and approved by a governing agency.

>>>>

	The GDSC database contains additional information on both the drugs and 
	molecular features, such as the target 
	pathways and developmental stages of the drugs. Additional online
	repositories may provide extra information such as the molecular weight of
	the compounds or the publication signatures of the molecular features. In some
	cases, prior knowledge on the drug efficacies may be available,
	from previous experiments. We propose to include these possibly beneficial 
	information sources in the estimation of the sensitivity prediction models in 
	a data-driven manner. More specifically, we estimate a normal 
	inverse Gaussian model, where the extent of regularisation is estimated by an 
	adaptive empirical Bayes procedure, guided by the external information.

@@

	The rest of the paper is structured as follows. In Section 
	\ref{sec:model} we introduce our model, the estimation of which is detailed
	in Section \ref{sec:estimation}. In Section \ref{sec:gdsc} we analyse the
	GDSC data, and we end with a discussion in Section \ref{sec:discussion} on the 
	pros and cons of the proposed method.

@@

	\subsection{Bayesian prior model}
	We carry out inference by endowing the parameters with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
		  \label{eq:tauprior}\\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\phi$ and shape $\lambda$. Prior distributions of the form
	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
	prior 
	variance into a local component $\gamma_{jd}^2$ and a global component
	$\tau_d^2$. The local components are supposed to capture local variation in
	the model parameters $\beta_{jd}$, while the global component is supposed to
	capture the general trend in $\bm{\beta}_{d}$. For appropriate 
	local shrinkage in global-local 
	shrinkage models it is important  to account for 
  different noise levels $\sigma_d^2$ by scaling the $\beta_{jd}$ variances 
  accordingly.

	The normal inverse Gaussian (NIG) prior model was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and since 
  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
  additional global variance component $\tau_d^2$. Supplementary Material
  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
  To illustrate the effect of the NIG prior on the posterior mean, we consider 
  the prior reparametrised as in 
  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
  Under the (simplified) normal means model, i.e., 
  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
  $\tau_d^2=\sigma_d^2=1$, the resulting conditional
  posterior mean for the $\beta_{jd}$ is
  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
  implies full shrinkage towards zero. Figure 
  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
  several choices of $\beta_{jd}$ prior.
<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
@

@@

	\subsection{External information}\label{sec:external}
	In drug sensitivity prediction models, external information on both the drugs
	and features is often available. Here, we assume this information to be 
	available as
	external feature `covariates' $\mathbf{c}_{jdg}$, for $g=1, \dots, G$, and 
	drug `covariates' $\mathbf{z}_{dh}$, for $h=1,\dots, H$. An example of a 
	(binary) feature covariate is target pathway presence, with $c_{jdg}=0$ if 
	gene $j$ is present in the 
	target pathway of drug $d$ and $c_{jdg}=1$ if it is not. An example of a 
	(ternary) drug
	covariate is developmental phase, with levels experimental phase, 
	clinical development, and approved by a governing agency.
